Gradual Soundness: Lessons from Static Python

### 2021-12-24 pj-7.1/paper.tex
pj: sec 5 rough draft, production experience

\newcommand{\typefont}[1]{\mathsf{#1}}
\newcommand{\paramtype}[2]{#1[#2]}
\newcommand{\sptype}{\typefont{T}}
\newcommand{\sptclass}{\typefont{C}}
\newcommand{\sptint}{\typefont{Int}}
\newcommand{\sptfloat}{\typefont{Float}}
\newcommand{\sptdyn}{\typefont{Dynamic}}
\newcommand{\sptobject}{\typefont{Object}}
\newcommand{\sptnone}{\typefont{NoneType}}
\newcommand{\sptinstanceof}[1]{\paramtype{\typefont{Instance}}{#1}}
\newcommand{\sptfun}[2]{#1 \rightarrow #2}
\newcommand{\sptunion}[2]{#1 \cup #2}
\newcommand{\sptoptional}[1]{\paramtype{\typefont{Optional}}{#1}}
\newcommand{\sptrawpydict}{\typefont{PyDict}}
\newcommand{\sptpydict}[2]{\paramtype{\sptrawpydict}{#1, #2}}
\newcommand{\sptchkdict}[2]{\paramtype{\typefont{ChkDict}}{#1, #2}}

@@

  \paragraph{Grounding}
  % What argument, feasibility proof, artifacts,
  % or results and evaluation support this work?
  Our model of \SP{} is implemented in PLT Redex and tested against
  the \SP{} regression suite.
  This paper includes a small core of the model to convey the main ideas
  of the \SP{} approach and its soundness.
  Our performance claims are supported by data from both
  a public suite a microbenchmarks and the proprietary Instagram codebase.

  \paragraph{Importance}
  % Why does this work matter?
  \SP{} is the first language that improves performance across the board through
  the addition of sound types to an untyped language.
  %% Prior work on Nom and Dart 2 showed that concrete types are promising.
  Other language designers may wish to replicate its approach,
  especially those who maintain unsound type checkers for untyped languages
  such as JavaScript.

@@

The \SP{} type system is a unique synthesis of ideas from the gradual typing literature
and prior work on types for Python.
Nom~\cite{mt-oopsla-2017,mt-oopsla-2021} and PEP 484~[CITE] are two notable sources.
Relative to these, the \SP{} takes some interesting turns
because of the engineering context at Instagram.
The following contextual points are important to bear in mind as we present
notable aspects of the type system:
\begin{itemize}
  \item
    Performance is the bottom line.
    Unless \SP{} helps production code run faster, the team will be disbanded
    and reallocated.
    %% 2021-12-14: too extreme? and what about soundness?
  \item
    %% Q. why is the python so nominal? style guide? java background?
    The codebase makes heavy use of first-order classes and objects.
    First-class classes, objects, and even functions are rare.
  \item
    Prior to investing in \SP{}, Instagram used Cython to compile
    performance-critical code to C extension modules.
    Cython performed well, but mixing pure Python with a few compiled modules
    made for an awkward developer experience.
    \SP{} aims to match the Python workflow.
  \item
    The Instagram codebase is already typed.
    Developers use Pyre [CITE], an optional type checker for Python,
    to write type annotations and to drive IDE tools (e.g., autocompletion).
\end{itemize}

@@

Three unsupported types bear special mention.
There are no first-class class or object types, as noted in \cref{s:inheritance}.
There are also no types for first-class functions (callables);
functions are common at Instagram, but first-class functions are rare.
Finally, \SP{} does not support recursive types due to low demand.
Neither Pyre nor mypy have recursive types either.
%% demand seems low ... baffling
%% mypy = https://github.com/python/mypy/issues/731
%% nothing in pep 484 = https://www.python.org/dev/peps/pep-0484/

@@

\SP{} does not attempt to type first-class classes:
partly because they see little use at Instagram (FILL how much?)
and partly because it is unclear how to incorporate them into the
nominal type system.
The straightforward but restrictive approach is to force code that uses a
first-class class to expect subtypes of a particular named static class.
Flatt et al.~\cite{fkf-popl-1998} propose a more-flexible approach, %(specifically for mixins)
but it requires a second layer of \emph{interface types} atop the nominal hierarchy.
MonNom~\cite{mt-oopsla-2021} uses interfaces in a similar way to accommodate
structural objects.
Adding an interface layer to \SP{} is an open question.

@@

\section{Evaluation}
\label{s:eval}

\subsection{Microbenchmarks}

richards etc.

%% https://github.com/facebookincubator/cinder/tree/cinder/3.8/Tools/benchmarks
%%
%% deltablue
%% fannkuch
%% nbody
%% nqueens
%% pystone
%% richards
%% richards_static
%%
%% UHOH except for richards_static, none of these have types!!!
%% Need to ask if C&D have other typed benchmarks around
%% ... if not, ask them to port these and measure and tune if needed.

\subsection{Production Experience}

%% 2021-12-15: C&D may have an SP off/on switch soon
%% - accumulated CPU savings since July 1 = ~2% total IG production CPU
%%   ... how does that number work?
%% - in the beginning (?) H1 got 1% from first conversions + 0.7% from all Cython (higher bar)
%%   - no Cython regressions, SP + JIT is good enough
%%   - some Cython still exists, imported as library code from another codebase
%% - rough measure: 0 to 25% per module
%%   ... but depends on nature of module
%%   ... doesn't measure the cooperative effect of static-to-static cross-module calls

%% - 530 SP modules
%%   - 400 from a codegen tool
%%   - 130 by hand
%%   - currently +~3/4 per week by hand

@@

%% - review of 30 winning diffs "provided the bulk of our wins this half"
%%   most common changes, not in careful order, are:
%%   1) adjustments to tests to pass correct types instead of mock objects, as as
%%      not to fail type checks in SP code. This is an inherent and desired
%%      incompatibility; next half we plan to work on improving the Python mock
%%      framework to automatically create mocks of the correct types.
%%      - bg: lookup the dino email about mocks
%%   2) adjustments to mock assertions in tests about calls to adjust for the
%%      fact that SP currently converts all call arguments to be positional (we plan
%%      to fix this incompatibility by falling back to the original call arguments
%%      when we detect the target function has been patched.)
%%      - bg: ??? bit fuzzy about positional and others
%%   3) adjustments around class attributes vs instance attributes (Python allows
%%      this to be a bit fuzzy, where a class attribute can be a fallback default
%%      for an instance attribute if not set, but SP currently requires an attribute
%%      to be clearly one or the other, and we have more efficient access for
%%      instance attributes via LOAD_FIELD that we don’t have for class attributes.)
%%      - related to github #37 ?
%%   4) convert @classmethod that don’t actually need the class to @staticmethod,
%%      allowing for emitting direct INVOKE_FUNCTION instead of INVOKE_METHOD
%%   5) extract special types that our compiler doesn’t yet support/understand
%%      (classes decorated with @dataclass or inheriting from `enum.Enum`) out of
%%      the module we are converting to static. We have work in progress on
%%      supporting some Enum types, and plan to add a @dataclass intrinsic soon.
%%   6) Removing the use of super-dynamic Python features like `__setattr__` that
%%      we don’t support. So far in the cases where we’ve done this it hasn’t
%%      required a major rewrite.
%%      - bg: more details pls ,,, sketch the rewrite
%%   7) Removing the use of keyword-only arguments (bare `*`) in function
%%      signatures, since we don’t yet support calling such functions. We plan to
%%      support this next half.
%%   8) Convert Python integers and bools used in very hot code paths into SP
%%      machine primitives.
%%      - bg: focus on block, convert, no problem?
%%   9) Occasionally converting a regular Python list or dict to a CheckedList or
%%      CheckedDict.
%%      - bg: how occasional? worried about power of the "gradual soundness" claim
%%
%% common theme: gotta change more tests than code
%%
%% significant change: replaced `contextlib.ContextDecorator` with `__static__.ContextDecorator`
%%  removes an extra call layer
%%  helps with (timing [ context managers / decorators ])
%%
%% all of this is the low-hanging fruit, haven't tackled uses of metaclasses even though looks promising
%%  because the pay / payoff isn't yet attractive

@@

>>>>

  \newcommand{\bmname}[1]{\textbf{#1}}
\newcommand{\typefont}[1]{\mathsf{#1}}
\newcommand{\paramtype}[2]{#1[#2]}
\newcommand{\sptype}{\typefont{T}}
\newcommand{\sptclass}{\typefont{C}}
\newcommand{\sptint}{\typefont{Int}}
\newcommand{\sptfloat}{\typefont{Float}}
\newcommand{\sptdyn}{\typefont{Dynamic}}
\newcommand{\sptobject}{\typefont{Object}}
\newcommand{\sptnone}{\typefont{NoneType}}
\newcommand{\sptinstanceof}[1]{\paramtype{\typefont{Instance}}{#1}}
\newcommand{\sptfun}[2]{#1 \rightarrow #2}
\newcommand{\sptunion}[2]{#1 \cup #2}
\newcommand{\sptoptional}[1]{\paramtype{\typefont{Optional}}{#1}}
\newcommand{\sptrawpydict}{\typefont{PyDict}}
\newcommand{\sptpydict}[2]{\paramtype{\sptrawpydict}{#1, #2}}
\newcommand{\sptchkdict}[2]{\paramtype{\typefont{ChkDict}}{#1, #2}}

@@

  \paragraph{Grounding}
  % What argument, feasibility proof, artifacts,
  % or results and evaluation support this work?
  Our model of \SP{} is implemented in PLT Redex and tested against
  the \SP{} regression suite.
  This paper includes a small core of the model to convey the main ideas
  of the \SP{} approach and its soundness.
  Our performance claims are based on production experience in the Instagram
  web server.
  Thanks to \SP{}, global CPU usage has improved by almost 4\% within a year.

  \paragraph{Importance}
  % Why does this work matter?
  \SP{} is the first sound gradual language whose piece-meal application
  to a realistic codebase has consistently improved performance.
  %% Prior work on Nom and Dart 2 showed that concrete types are promising.
  Other language designers may wish to replicate its approach,
  especially those who maintain unsound type checkers for untyped languages
  such as JavaScript.

@@

The \SP{} type system is a unique synthesis of ideas from the gradual typing literature
and prior work on types for Python.
Nom~\cite{mt-oopsla-2017,mt-oopsla-2021} and PEP 484~[CITE] are two notable sources.
Relative to these, the \SP{} takes some interesting turns
because of the engineering context at Instagram.
The following contextual points are important to bear in mind as we present
notable aspects of the type system:
\begin{itemize}
  \item
    Performance is the bottom line.
    Unless \SP{} helps production code run faster, the team will be disbanded
    and reallocated.
    %% 2021-12-14: too extreme? and what about soundness?
  \item
    %% Q. why is the python so nominal? style guide? java background?
    The codebase depends heavily on first-order functions and objects in
    performance-critical code.
    Higher-order code is fairly common, but rarely appears on a critical path.
  \item
    Prior to investing in \SP{}, Instagram used Cython to compile
    performance-critical code to C extension modules.
    Cython performed well, but mixing pure Python with a few compiled modules
    made for an awkward developer experience.
    \SP{} aims to match the Python workflow.
  \item
    The Instagram codebase is already typed.
    Developers use Pyre [CITE], an optional type checker for Python,
    to write type annotations and to drive IDE tools (e.g., autocompletion).
\end{itemize}

@@

Three unsupported types bear special mention.
There are no first-class class or object types, as noted in \cref{s:inheritance}.
There are also no types for first-class functions (callables).
Pyre can check these types instead, to point out clear logical flaws.
Finally, \SP{} does not support recursive types due to low demand.
Neither Pyre nor mypy have recursive types either.
%% demand seems low ... baffling
%% mypy = https://github.com/python/mypy/issues/731
%% nothing in pep 484 = https://www.python.org/dev/peps/pep-0484/

@@

\SP{} does not attempt to type first-class classes: partly because they have
yet to appear in performance-critical code at Instagram and partly because it
is unclear how to incorporate them into the nominal type system.
The straightforward but restrictive approach is to force code that uses a
first-class class to expect subtypes of a particular named static class.
Flatt et al.~\cite{fkf-popl-1998} propose a more-flexible approach, %(specifically for mixins)
but it requires a second layer of \emph{interface types} atop the nominal hierarchy.
MonNom~\cite{mt-oopsla-2021} uses interfaces in a similar way to accommodate
structural objects.
Adding an interface layer to \SP{} is an open question.

@@

\section{Production Experience}
\label{s:eval}
%% 2021-12-15: C&D may have an SP off/on switch soon to measure totals

% "The workload for those rough percentage numbers is “Instagram web server
%  production.” We have pretty extensive production profiling that lets us
%  identify CPU savings with good granularity when a change lands in prod, and the
%  observed CPU savings for a variety of staticification changes we’ve landed over
%  the last half sum up to 2% of total IG prod CPU usage." --Carl

\begin{table}
  \caption{Overview: \SP{} in production}
  \label{t:prod-stat}
  \begin{tabular}{l@{~~}ll@{~~}l}
    8     & months in production & 3.7\% & global CPU improvement \\
    530   & modules converted  &  0 & performance regressions
    %% TODO 530 / ???? how much of codebase?
  \end{tabular}
\end{table}

The Instagram web server has been running \SP{} in production since April 2021.
Overall, the results are very encouraging.
Instagram's internal profiling tools, which continuously monitor
performance, attribute a 3.7\% improvement in global CPU to \SP{} conversions.
\cref{t:prod-stat} reports a few other production statistics.
Over 500 modules have been converted thus far.
None of the converted modules have gotten slower because of \SP{}

\subsection{The Migration Path}

\SP{} first entered the Instagram web server as a replacement for Cython,
which Instagram had been using to speed up critical modules.
Cython performed well, but its partial adoption led to an awkward workflow
because unlike standard Python, Cython is a compiled language.
Replacing these modules with \SP{} let developers return to a conventional
Python workflow.
\SP{} (combined with Cinder) did not slow down any of these
performance-critical modules, and even sped a few up.
The net improvement was a 0.7\% reduction in global CPU usage.

Later \SP{} migrations have been led by profiling to find frequently-executed code.
The goal of these opportunistic migrations was to test whether \SP{} could
improve typical modules without slowing down their neighbors.
As a point of comparison, Typed Racket can improve individual Racket
modules but the type checks that it requires on module boundaries
can lead to huge slowdowns~\cite{tfgnvf-popl-2016,gtnffvf-jfp-2019}.
\SP{} does not appear to suffer in the same way.

During the first half of 2021, the \SP{} team identified critical
modules and proposed types to the maintainers of these modules.
Often, the types came with small code changes.
The accepted types and changes resulted in a 1\% improvement
in global CPU usage.
During the second half of 2021, the \SP{} team applied the same
process at a larger scale by incorporating a tool that drafts
initial type annotations.
These changes resulted in a 2\% global CPU improvement.

As of December 2021, the \SP{} team has converted 530 modules.
Most of these came from the annotation tool (400); the rest
are from hand conversions.
In a typical week, the team adds types to four additional modules.
These modules have seen speedups ranging from 0\% to 25\% depending
on their nature.
Regressions have not been an issue.
The only cause for alarm is when a module gets only a small improvement
without code changes.
%% - rough measure: 0 to 25% per module
%%   ... but depends on nature of module
%%   ... doesn't measure the cooperative effect of static-to-static cross-module calls

\subsection{Analyzing Code Changes}

The fact that the \SP{} team has been changing code as well as types
during its migrations is potentially troubling.
It could be that the \SP{} approach to gradual typing only pays off
after significant modifications, or only for a restrictive style of programming.

To understand the code changes, an author affiliated with Instagram reviewed
approximately 30 patches that significantly improved performance.
In general, the patches contain relatively minor code changes.
The biggest changes are to testing code rather than production code.
Both these points are encouranging.
The following nine points describe common patterns:

\begin{enumerate}
  \item
   adjustments to tests to pass correct types instead of mock objects, as as
   not to fail type checks in SP code. This is an inherent and desired
   incompatibility; next half we plan to work on improving the Python mock
   framework to automatically create mocks of the correct types.
   %% - bg: lookup the dino email about mocks

 \item
   adjustments to mock assertions in tests about calls to adjust for the
   fact that SP currently converts all call arguments to be positional (we plan
   to fix this incompatibility by falling back to the original call arguments
   when we detect the target function has been patched.)
   %% - bg: ??? bit fuzzy about positional and others

 \item
   adjustments around class attributes vs instance attributes (Python allows
   this to be a bit fuzzy, where a class attribute can be a fallback default
   for an instance attribute if not set, but SP currently requires an attribute
   to be clearly one or the other, and we have more efficient access for
    instance attributes via \code{LOAD\_FIELD} that we don’t have for class attributes.)
   %% - related to github #37 ?

 \item
   convert @classmethod that don’t actually need the class to @staticmethod,
    allowing for emitting direct \code{INVOKE\_FUNCTION} instead of \code{INVOKE\_METHOD}

 \item
   extract special types that our compiler doesn’t yet support/understand
   (classes decorated with @dataclass or inheriting from `enum.Enum`) out of
   the module we are converting to static. We have work in progress on
   supporting some Enum types, and plan to add a @dataclass intrinsic soon.

 \item
   Removing the use of super-dynamic Python features like \code{\_\_setattr\_\_} that
   we don’t support. So far in the cases where we’ve done this it hasn’t
   required a major rewrite.
   %% - bg: more details pls ,,, sketch the rewrite

 \item
   Removing the use of keyword-only arguments (bare `*`) in function
   signatures, since we don’t yet support calling such functions. We plan to
   support this next half.

 \item
   Convert Python integers and bools used in very hot code paths into SP
   machine primitives.
   %% - bg: focus on block, convert, no problem?

 \item
   Occasionally converting a regular Python list or dict to a CheckedList or
   CheckedDict.
   %% - bg: how occasional? worried about power of the "gradual soundness" claim
\end{enumerate}

%% significant change: replaced `contextlib.ContextDecorator` with `__static__.ContextDecorator`
%%  removes an extra call layer
%%  helps with (timing [ context managers / decorators ])
%%
%% all of this is the low-hanging fruit, haven't tackled uses of metaclasses even though looks promising
%%  because the pay / payoff isn't yet attractive

\subsection{Developer Experience}

@@

%% - pyre challenges?

\subsection{Microbenchmarks}
%% https://github.com/facebookincubator/cinder/tree/cinder/3.8/Tools/benchmarks

%% TODO jit vs nojit

%% 2021-12-23: TODO decide whether to ask for typed benchmarks to report ratios
%%  Carl: we could commit to typed 2--3 more, but pystone etc (?!) are notoriously bad

%% Somehow though, we ought to have public code available.

Prior to deploying \SP{} in production, the team used microbenchmarks to study performance.
These microbenchmarks are admittedly small and quite different than typical application code.
That said, their use gave the team a coarse idea of whether \SP{} was on track during development.
We include these microbenchmarks to offer reproducible performance numbers;
unlike the Instagram web server, the benchmark code is public.

\Cref{t:microbenchmark} reports current microbenchmark results.
Each row gives the ratio of two running times: fully-typed \SP{} code versus
unannotated \SP{} code.
A ratio between 0 and 1 means that performance improved thanks to the types.
[FILL] numbers.

\begin{table}[t]
  \caption{Microbenchmark Performance Ratios}
  \label{t:microbenchmark}
  %% show untyped Python time?
  \begin{tabular}{ll}
    Name               & Typed / Dynamic \\\midrule
    \bmname{deltablue} &             0\% \\
    \bmname{fannkuch}  &             0\% \\
    \bmname{nbody}     &             0\% \\
    \bmname{nqueens}   &             0\% \\
    \bmname{pystone}   &             0\% \\
    \bmname{richards}  &         0.125\% \\
  \end{tabular}
\end{table}

%% TODO conclusions? threats?

@@

Grace nominal transient aint bad at all
